### 2025-05-27 lecturenotes_aima203/ch2_aima203.tex
ch2

\section{Iteration Method}
Steps:
\begin{enumerate}
\item Re-write the given equation \(f(x)=0\) in the form \(x= \phi (x)\).
  This equation is of \textbf{iterative-type}. Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\), if the iteration is of convergent one.
\item Choose an initial root of \(f\), \(x_0\).
\item \(x_1=\phi (x_0)\), \(x_2= \phi(x_1)\) and so on.
\end{enumerate}
The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number. But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation.

\subsection{Exercise}
Using Iteration Method:
\begin{enumerate}
\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places.
\end{enumerate}

@@

\item Let \(x_1\) be a solution, which is more close to the exact solution of \(f(x)=0\). Then Using Taylor's expansion of \(f\) about \(x_0\): \[f(x_1)=f(x_0) + (x_1-x_0)f'(x_0) + (x_1+x_0)^2f''(x_0) + ... = 0\]
  Neglecting the second and higher order derivatives, we get
  \begin{equation}
    \label{taylor}
  f(x_0) + (x_1-x_0)f'(x_0)=0
  \end{equation}
  \begin{footnotesize}
    The equation-\ref{taylor} is a linear equation, so this is an linear approximation. This equation is infact the tangent to the curve of the function \(f(x)\) at $(x_0, f(x_0))$. And it is the point $x_1$ where the tangent meets the $x-axis$. So, the next approximation after $x_0$ by Newton-Rapshon's method is the point on $x-axis$, where the tangent to the $f$ at $x_0$ meets the $x-axis$. This point can be solved as follows:
   \end{footnotesize}
  \begin{align*}
    x_1-x_0 & = - \frac{f(x_0)}{f'(x_0)} \\
    x_1 &= x_0 - \frac{f(x_0)}{f'(x_0)}
  \end{align*}

@@

\subsection{Exercise}
Using Newton-Rapshon's Method:
\begin{enumerate}
\item Find a root of \(f(x)=xe^x-1=0\), correct to \(4\) decimal places.
\item Find a non-zero root of the equation $x^2 + 4sinx = 0$.
\item Using the Newton-Raphson method, derive a formula for finding the $kth$ root of a positive number $N$ and hence compute the value of $(25)^{1/4}$.

@@

\subsubsection{Exercise}
\begin{enumerate}
\item Find a real root of the equation $x^3-2x-5=0$.
\end{enumerate}

>>>>

\section{Iteration Method}
Steps:
\begin{enumerate}
\item Re-write the given equation \(f(x)=0\) in the form \(x= \phi (x)\).
  This equation is of \textbf{iterative-type}. Meaning we can substitute a value of \(x\) in \(\phi (x)\) to get another value of \(x\), and continue this process to get the desired value of \(x\), if the iteration is of convergent one.
\item Choose an initial root of \(f\), \(x_0\).
\item \(x_1=\phi (x_0)\), \(x_2= \phi(x_1)\) and so on.
\end{enumerate}
The sequence \(x_0, x_1, x_2, ...\) may not converge to a definite number. But if the sequence converges to a definite number \(\zeta\), then \(\zeta\) is a root of the given equation. The sufficient condition for the sequence of the approximations $x_0, x_1, x_2,...$ by iteration method to converge is that $|\phi ' (x) | < 1$ in some neighborhood (interval) of $x_0$.

\subsection{Exercise}
Using Iteration Method:
\begin{enumerate}
\item Find a root of \(2x-3-cosx=0\), correct to \(3\) decimal places lying in $[3/2, \pi/2]$.
\end{enumerate}

@@

\item Let \(x_1\) be a solution, which is more close to the exact solution of \(f(x)=0\). Then Using Taylor's expansion of \(f\) about \(x_0\): \[f(x_1)=f(x_0) + (x_1-x_0)f'(x_0) + (x_1-x_0)^2f''(x_0) + ... = 0\]
  Neglecting the second and higher order derivatives, we get
  \begin{equation}
    \label{taylor}
  f(x_0) + (x_1-x_0)f'(x_0)=0
  \end{equation}
  \begin{footnotesize}
    The equation-\ref{taylor} is a linear equation, so this is an linear approximation. This equation is infact the tangent to the curve of the function \(f(x)\) at $(x_0, f(x_0))$. And it is the point $x_1$ where the tangent meets the $x-axis$. So, the next approximation after $x_0$ by Newton-Rapshon's method is the point on $x-axis$, where the tangent to the $f$ at $x_0$ meets the $x-axis$. This point can be solved as follows:
   \end{footnotesize}
  \begin{align*}
    x_1-x_0 & = - \frac{f(x_0)}{f'(x_0)} \\
    x_1 &= x_0 - \frac{f(x_0)}{f'(x_0)}
  \end{align*}

@@

\subsection{Exercise}
Using Newton-Rapshon's Method:
\begin{enumerate}
\item Find a root of \(f(x)=xe^x-1=0\), correct to \(4\) decimal places.
\item Find a non-zero root of the equation $x^2 + 4sinx = 0$ choosing $x_0 = \pi$.
\item Using the Newton-Raphson method, derive a formula for finding the $kth$ root of a positive number $N$ and hence compute the value of $(25)^{1/4}$.

@@

\subsubsection{Exercise}
\begin{enumerate}
\item Find a real root of the equation $x^3-2x-5=0$ with three iterations.
\end{enumerate}

### 2025-05-27 lecturenotes_aima203/ch3_aima.tex
ch2

\subsection{Tridiagonal Systems}
The matrix of the form $$A= \begin{bmatrix}
    a_{11} & a_{12} & 0 & 0 &... & 0 & 0 & 0 & 0\\
    a_{21} & a_{22} & a_{23} & 0 &... & 0 &0 &0 & 0\\
    0 & a_{32} & a_{33} & a_{34} &... & 0 & 0 &0 & 0\\
    ...\\
    0 & 0 & 0 & 0 &...& 0 & a_{(n-1)(n-3)} & a_{(n-1)(n-2)} & a_{(n-1)(n-1)} \\
    0 & 0 & 0 & 0&... & 0 & 0 & a_{n(n-2)} & a_{n(n-1)}
\end{bmatrix}$$ is called tridiagonal matrix.
The method of $LU$ factorization can conveniently apllied to the system having a tridiagonal matrix as the coefficient matirx.

\subsection{Exercise}
Solve the system:
\begin{enumerate}
\item $2x-y=0, \;\; -x+2y-z=0, \;\; -y+2z-u=0, \;\; -z+2u=1$.
\item $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$
\end{enumerate}
\section{Vector and Matrix Norms}
\textbf{Vector Norm} \\[1mm]
The distance between a vector and the null vector is a measure of the \textit{length} of the vector. This is called a norm of the vector. Mathematically, a norm of a vector in a vector space $V$ is defined as a function $\displaystyle \Vert .  \Vert : V \mapsto \mathbb{R}$ that satisfies the following conditions:
\begin{enumerate}
\item [i)] $\Vert x \Vert \geq 0$
\item [ii)] $\Vert x \Vert = 0$ if and only if $x=0$.
\item [iii)] $\alpha \Vert x \Vert = \vert \alpha \vert \Vert x \Vert $ for any real $\alpha$.
\item [iv)] $\Vert x + y \Vert \leq \Vert x \Vert + \Vert y \Vert$.
\end{enumerate}
There are different types of length of a vector. These gives rise to different types of norms. Some useful norms are as follows:
\begin{enumerate}
\item $L1$ norm also called \\
  \textbf{Rectangular norm or Manhattan norm}: $\Vert x \Vert _1 = |x_1| + |x_2| +... |x_n| $.
\item $L2$ norm also called \textbf{Euclidean norm}: $\Vert x \Vert _2 = \sqrt{|x_1|^2 + |x_2|^2 +... |x_n|^2} $.
\item $L\infty$ norm also called \\
  \textbf{Maximum norm or Uniform norm}: $\Vert x \Vert _{\infty} = max\{|x_1| , |x_2| ,..., |x_n|\} $.
\end{enumerate}
\noindent
\textbf{Matrix Norm} \\[1mm]
Corresponding to the vector norms we have the three matrix norms defned as follows:
\begin{enumerate}
\item \textbf{Column Norm}: $\displaystyle \Vert A \Vert _1 = \max_j \sum_i |a_{ij}|$.
\item \textbf{Euclidean Norm}: $\displaystyle \Vert A \Vert _e = \left [\sum_{i,j} |a_{ij}|^2 \right ] ^{1/2}$.
\item \textbf{Column Norm}: $\displaystyle \Vert A \Vert _{\infty} = \max_i \sum_j |a_{ij}|$.
\end{enumerate}
\subsection{Exercise}
\begin{enumerate}
\item Find $L1, \; L2, \; L\infty$ norms of $x=(1,\,-13,\,5,\,3,\,-4$.
\item Find the Column, the Euclidean and the Row norm of $A =
  \begin{bmatrix}
    1 & 2 & 3 \\
    4 & 5 & 6 \\
    7 & 8 & 9
  \end{bmatrix}
  $
\end{enumerate}

@@

\begin{definition}[Conition Number]
  The quantity $c(A) = \Vert A \Vert \Vert A^{-1} \Vert$, gives the measure of the condition of the matrix A, and is called the condition number. Here $\Vert . \Vert$ is any matrix norm.
\end{definition}
Ill-conditioning can usually be expected when $|A|$, in the system $Ax=b$, is small. Let $A=[a_{ij}]$ and $\displaystyle s_i = \left [ a^2_{i1} + a^2_{i2} + ... + a^2_{in} \right ] ^{1/2}$. Let $\displaystyle k = \frac{|A|}{s_1s_2...s_n}$. Then the system whose coefficient matrix is $A$ is ill-conditioned if $k$ is very small compared to unity. Otherwise, it is well-conditioned.

@@

\section{Iterative Method}
The general system of linear equations with $n$ unknowns and $n$ linear equations is:
\begin{gather*}
    a_{11}x_1+a_{12}x_2+...+a_{1n}x_n=b_1 \\
    a_{21}x_1+a_{22}x_2+...+a_{2n}x_n=b_2 \\
    .\\
    .\\
    .\\
    a_{n1}x_1+a_{n2}x_2+...+a_{nn}x_n=b_n
  \end{gather*}
  For the iterative this system has to be rearranged as following
  \begin{align*}
    x_1 &= \frac{b_1}{a_{11}} - \frac{a_{12}}{a_{11}}x_2-\frac{a_{13}}{a_{11}}x_3-...-\frac{a_{1n}}{a_{11}}x_n \\
    x_2 &= \frac{b_2}{a_{22}} -\frac{a_{21}}{a_{22}}x_1-\frac{a_{23}}{a_{22}}x_3 ...-\frac{a_{2n}}{a_{22}}x_n \\
    .\\
    .\\
    .\\
    x_n &= \frac{b_n}{a_{nn}} - \frac{a_{n1}}{a_{nn}}x_1-\frac{a_{n2}}{a_{nn}}x_2-...-\frac{a_{n(n-1)}}{a_{nn}}x_{n-1}
  \end{align*}
  We can write this system as $x=Bx+C$. This equation is a iterative in nature as $x$ is a function of $x$-itself.
  So the iteration formula is
  \begin{equation}
    \label{simul}
    x^{n+1} = Bx^{n} + C
  \end{equation}
  \begin{enumerate}
  \item \textbf{Method of simultaneous displacements} \\[1mm]
    This method~\ref{simul} is due to \textit{Jacobi} and is called method of simultaneous displacements, because $x_1, ..., x_n$ values are displaced simultaneous in \ref{simul}. It can be shown that a sufficient condition for the convergence of this method is $\Vert B \vert < 1$.
 \item \textbf{Method of successive displacements} But if the values $x_1,...,x_n$ are displaced sucessively; obtain $x_1$ first from the initialization, then use this value to obtain $x_2$ and so on, then this medthod is called method of successive displacements which is also commonly known as \textit{Gauss-Seidel method}.
 \end{enumerate}

 \subsection{Exercise}
 \begin{enumerate}
 \item $10x-2y-z-u=3, \;\;\; -2x+10y-z-u = 15, \;\;\; -x-y+10z-2u=27, \\ -x-y-2z+10u = 9$
   \item $2x-y=1, \;\;\; -x+3y-z = 8, \;\;\; -y+2z=-5$
 \end{enumerate}

@@

>>>>

\subsection{Tridiagonal Systems}
The matrix of the form $$A= \begin{bmatrix}
    a_{11} & a_{12} & 0 & 0 &... & 0 & 0 & 0 & 0\\
    a_{21} & a_{22} & a_{23} & 0 &... & 0 &0 &0 & 0\\
    0 & a_{32} & a_{33} & a_{34} &... & 0 & 0 &0 & 0\\
    ...\\
    0 & 0 & 0 & 0 &...& 0 & a_{(n-1)(n-2)} & a_{(n-1)(n-1)} & a_{(n-1)n} \\
    0 & 0 & 0 & 0&... & 0 & 0 & a_{n(n-1)} & a_{nn}
\end{bmatrix}$$ is called tridiagonal matrix. It is a square matrix with nonzero elements only on the main diagonal, subdiagonal, superdiagonal.
The method of $LU$ factorization can conveniently apllied to the system having a tridiagonal matrix as the coefficient matirx.

\subsection{Exercise}
Solve the system:
\begin{enumerate}
\item $2x-y=0, \;\; -x+2y-z=0, \;\; -y+2z-u=0, \;\; -z+2u=1$.
\item $3x_1 - x_2 = -1, \;\;\; -x_1+3x_2-x_3 = 7, \;\;\; -x_2+3x_3 =7$
\end{enumerate}
\section{Vector and Matrix Norms}
\textbf{Vector Norm} \\[1mm]
The distance between a vector and the null vector is a measure of the \textit{length} of the vector. This is called a norm of the vector. Mathematically, a norm of a vector in a vector space $V$ is defined as a function $\displaystyle \Vert .  \Vert : V \mapsto \mathbb{R}$ that satisfies the following conditions:
\begin{enumerate}
\item [i)] $\Vert x \Vert \geq 0$
\item [ii)] $\Vert x \Vert = 0$ if and only if $x=0$.
\item [iii)] $\alpha \Vert x \Vert = \vert \alpha \vert \Vert x \Vert $ for any real $\alpha$.
\item [iv)] $\Vert x + y \Vert \leq \Vert x \Vert + \Vert y \Vert$.
\end{enumerate}
There are different types of length of a vector. These gives rise to different types of norms. Some useful norms are as follows:
\begin{enumerate}
\item $L1$ norm also called \\
  \textbf{Rectangular norm or Manhattan norm}: $\Vert x \Vert _1 = |x_1| + |x_2| +... |x_n| $.
\item $L2$ norm also called \textbf{Euclidean norm}: $\Vert x \Vert _2 = \sqrt{|x_1|^2 + |x_2|^2 +... |x_n|^2} $.
\item $L\infty$ norm also called \\
  \textbf{Maximum norm or Uniform norm}: $\Vert x \Vert _{\infty} = max\{|x_1| , |x_2| ,..., |x_n|\} $.
\end{enumerate}
\noindent
\textbf{Matrix Norm} \\[1mm]
Corresponding to the vector norms we have the three matrix norms defned as follows:
\begin{enumerate}
\item \textbf{Column Norm}: $\displaystyle \Vert A \Vert _1 = \max_j \sum_i |a_{ij}|$.
\item \textbf{Euclidean Norm}: $\displaystyle \Vert A \Vert _e = \left [\sum_{i,j} |a_{ij}|^2 \right ] ^{1/2}$.
\item \textbf{Column Norm}: $\displaystyle \Vert A \Vert _{\infty} = \max_i \sum_j |a_{ij}|$.
\end{enumerate}
\subsection{Exercise}
\begin{enumerate}
\item Find $L1, \; L2, \; L\infty$ norms of $x=(1,\,-13,\,5,\,3,\,-4$.
\item Find the Column, the Euclidean and the Row norm of $A =
  \begin{bmatrix}
    1 & 3 & 5 \\
    1 & 4 & 3 \\
    1 & 3 & 2
  \end{bmatrix}
  $
\end{enumerate}

@@

\begin{definition}[Condition Number]
  The quantity $c(A) = \Vert A \Vert \Vert A^{-1} \Vert$, gives the measure of the condition of the matrix A, and is called the condition number. Here $\Vert . \Vert$ is any matrix norm.
\end{definition}
Ill-conditioning can usually be expected when $|A|$, in the system $Ax=b$, is small. Let $A=[a_{ij}]$ and $\displaystyle s_i = \left [ a^2_{i1} + a^2_{i2} + ... + a^2_{in} \right ] ^{1/2}$. Let $\displaystyle k = \frac{|A|}{s_1s_2...s_n}$. Then the system whose coefficient matrix is $A$ is ill-conditioned if $k$ is very small compared to unity. Otherwise, it is well-conditioned.

@@

\section{Iterative Method}
The general system of linear equations with $n$ unknowns and $n$ linear equations is:
\begin{gather*}
    a_{11}x_1+a_{12}x_2+...+a_{1n}x_n=b_1 \\
    a_{21}x_1+a_{22}x_2+...+a_{2n}x_n=b_2 \\
    .\\
    .\\
    .\\
    a_{n1}x_1+a_{n2}x_2+...+a_{nn}x_n=b_n
  \end{gather*}
  For the iterative this system has to be rearranged as following
  \begin{align*}
    x_1 &= \frac{b_1}{a_{11}} - \frac{a_{12}}{a_{11}}x_2-\frac{a_{13}}{a_{11}}x_3-...-\frac{a_{1n}}{a_{11}}x_n \\
    x_2 &= \frac{b_2}{a_{22}} -\frac{a_{21}}{a_{22}}x_1-\frac{a_{23}}{a_{22}}x_3 ...-\frac{a_{2n}}{a_{22}}x_n \\
    .\\
    .\\
    .\\
    x_n &= \frac{b_n}{a_{nn}} - \frac{a_{n1}}{a_{nn}}x_1-\frac{a_{n2}}{a_{nn}}x_2-...-\frac{a_{n(n-1)}}{a_{nn}}x_{n-1}
  \end{align*}
  We can write this system as $x=Bx+C$. This equation is a iterative in nature as $x$ is a function of $x$-itself.
  So the iteration formula is
  \begin{equation}
    \label{simul}
    x^{n+1} = Bx^{n} + C
  \end{equation}
  \begin{mdframed}
    The necessary condition for the convergence of the iterative method is that the diagonal elements $a_{ii}$ do not vanish. If this is not the case then, the equations should be rearranged so that, the system satisfies this condition.
  \end{mdframed}

  \begin{enumerate}
  \item \textbf{Method of simultaneous displacements} \\[1mm]
    This method~\ref{simul} is due to \textit{Jacobi} and is called method of simultaneous displacements, because $x_1, ..., x_n$ values are displaced simultaneous in \ref{simul}. It can be shown that a sufficient condition for the convergence of this method is $\Vert B \Vert < 1$.
 \item \textbf{Method of successive displacements} But if the values $x_1,...,x_n$ are displaced sucessively; obtain $x_1$ first from the initialization, then use this value to obtain $x_2$ and so on, then this medthod is called method of successive displacements which is also commonly known as \textit{Gauss-Seidel method}.
 \end{enumerate}

 \begin{mdframed}
   The Jacobi and Gauss-Seidel methods converge for any choice of the first approximation $(x_j^{(1)})$ $j=1,2,...,n$, if every equation of the system satisfies the condition that
   \begin{equation}
     \label{converge}
     \sum ^{n}_{j=1, j \neq i} \; \left \vert \frac{a_{ij}}{a_{ii}} \right \vert \leq 1, \hspace{7mm} i=1,2,...,n
   \end{equation}
   It is found that the Gauss-Seidel method converges \textbf{twice as fast} as the Jacobi method.
 \end{mdframed}
 \subsection{Exercise}
 \begin{enumerate}
 \item $10x-2y-z-u=3, \;\;\; -2x+10y-z-u = 15, \;\;\; -x-y+10z-2u=27, \\ -x-y-2z+10u = 9$
 \item $2x-y=1, \;\;\; -x+3y-z = 8, \;\;\; -y+2z=-5$

 \item An approximate solution of the system $10x_1 + x_2 +x_3 = 12, \hspace{3mm} x_1 + 10x_2 +x_3 = 12, \hspace{3mm} x_1 + x_2 + 10x_3 =12$ is given as $x_1^{(0)}=0.4, \hspace{3mm} x_2^{(0)}=0.6, \hspace{3mm} x_3^{(0)}=0.8$. Use the iterative method to improve this solution.

 \item Solve the system: $10x + 2y +z = 9, \hspace{3mm} 2x + 20y -2z = -44, \hspace{3mm} -2x+3y +10z = 22$.

 \item Apply upto six iterations of Gauss-Seidel to solve: $28x+4y-z=32, \hspace{3mm} 2x+17y+4z=35, \hspace{3mm} x+3y+10z=24$.

 \item \textbf{Cholesky's method} A matrix $A$ is called a symmetric matrix if $a_{ij} =a_{ji}$. For symmetric matrix the $LU$ decomposition is more conveniently obtained, since $U=L^{T}$. This method is called Cholesky's method. Solve the system by cholesky's method : $5x_1 + x_3 = 8 , \hspace{3mm} -2x_2 =-4, \hspace{3mm} x_1 + 5x_3 = 16$
 \end{enumerate}

@@

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:
